The finite representation property for some reducts of relation algebras

### 2021-04-17 frp-paper/Paper.tex
elementary chain argument for uncountable join semilattice-ordered semigroups

@@

\begin{lemma} Let $\mathcal{A} = \langle A, ;, + \rangle$ be a join semilattice-ordered semigroup,

  \begin{enumerate}
    \item If $\mathcal{A}$ is representable then $\exists$ has a winning strategy in $\mathcal{G}_{\omega}(\mathcal{A})$.
    \item If $\mathcal{A}$ is infinite and $\exists$ has a winning strategy in $\mathcal{G}_{\omega}(\mathcal{A})$ then $\mathcal{A}$ is representable.
  \end{enumerate}
\end{lemma}

@@

    \item
    Firstly, let us suppose that $|\mathcal{A}| = \omega$.

    Let $a, b \in \mathcal{A}$ such that $a \not\leq b$. We consider a play of the game where $\forall$ pick the pair $(a, b)$ in the initial round and plays $(\mathcal{N}, x, y, z, c, d)$ in the further rounds for each $x, y, z \in \operatorname{Nodes}(\mathcal{N})$ and $c,d \in \mathcal{A}$. Here, $c \in l_N(x, y)$ and $d \in l_N(y, z)$.

@@

\end{itemize}
For $|\mathcal{A}| \geq \aleph_1$ we apply the previous item and Tarksi-Vaught theorem on elementary chain union (see \cite[Theorem 2.5.2]{hodges1997shorter}).
  \end{enumerate}
\end{proof}

@@

\begin{proposition}
  Let $\mathcal{A}$ be a join semilattice-ordered semigroup, then there exists a game such that $\mathcal{A}$ is representable iff $\exists$ has a winning strategy in the all finite lenght games.
\end{proposition}

@@

@@

Let $\mathcal{A}$ be a join semilattice, a subset $I \subseteq \mathcal{A}$ is called \emph{ideal}, if the following hold:
\begin{enumerate}
  \item If $x \in I$, then $\down{x} \subseteq I$,
  \item $x, y \in I$ implies $x + y \in I$.
\end{enumerate}
The poset $\langle \operatorname{Id}(\mathcal{A}), \subseteq \rangle$ consists of all ideals of $\mathcal{A}$ ordered by inclusion. A join semilattice $\mathcal{A}$ is called \emph{distributive}, if for each $x, a, b \in \mathcal{A}$ such that $x \leq a + b$, so there are $a' \leq a$ and $b' \leq b$ such that $x = a' + b'$. A finite $\mathcal{A}$ is distributive is the lattice $\operatorname{Id}(\mathcal{A})$ is distributive in the usual sense, see \cite[Lemma 184]{gratzer2011lattice}.

Note that an arbitrary finite $\mathcal{A} \in {\bf R}(;,+)$ does not have to be a quantale since a finite join semilattice may not have the bottom. Let us extend $\mathcal{A}$ with the bottom defined as $0 = \Sigma \emptyset$, we denote this algebra as $\mathcal{A}_0$. Note that $a_1 ; a_2 = 0$, where $a_1, a_2 \in \mathcal{A}_0$ and one has either $a_1 = 0$ or $a_2 = 0$.

\begin{proposition}
  If $\mathcal{A} \in {\bf R}(;,+)$ is finite, then $\mathcal{A}_0$ is representable.
\end{proposition}

\begin{proof}
  Let $h$ be a representation of $\mathcal{A}$ over some base set $D$. Let us define a map $h_0 : \mathcal{A}_0 \to 2^{D \times D}$ such that $h_0(a) = h(a)$, if $a \neq 0$, and $h_0(0) = \emptyset$. $h_0$ is a representation since $h$ is a representation.
\end{proof}

\begin{proposition} \label{distr}
  Let $\mathcal{A} \in {\bf R}(;,+)$ and $\mathcal{A}$ is finite, then
  \begin{enumerate}
    \item $\mathcal{A}_0$ is distributive,
    \item Every generator of $\mathcal{A}_0$ as a quantale is join-prime.
  \end{enumerate}
\end{proposition}

  \begin{enumerate}
    \item
    $\mathcal{A}$ is finite, so the $\operatorname{Id}(\mathcal{A}_0)$ consists of principal ideals, that is, lower cones generated by elements of $\mathcal{A}$.

    \begin{center}
    $\operatorname{Id}(\mathcal{A}_0) = \{ \down{a} \: | \: a \in \mathcal{A} \}$
    \end{center}

    Let $h$ be a representation of $\mathcal{A}$, consider $h_0(\mathcal{A}_0)$ and its set of ideals $\operatorname{Id}(h_0(\mathcal{A}_0))$. Intersection of ideals is defined is usual since intersection of ideals is an ideal, join is defined as follows:

    \begin{center}
    $\down{h_0(a)} \vee \down{h_0(b)} = \down{(h_0(a) \cup h_0(b))}$
    \end{center}
    Note that $\down{(h_0(a) \cup h_0(b))} = \down{h_0(a + b)}$ since $h$ is a representation of $\mathcal{A}$.

    Let $c \in \mathcal{A}$, consider $h(c)$.

    Suppose $h(d) \in \down{h_0(c)} \cap \down{h_0(a + b)}$. We need $h_0(d) \in \down{(h_0(c) \cap h_0(a))} \vee \down{(h_0(c) \cap h_0(b))}$.

    Let $(x, y) \in h_0(d)$, then either $(x, y) \in h_0(a)$ or $(x, y) \in h_0(b)$ since $d \leq a + b$.
    Suppose $(x, y) \in h_0(a)$. We need $h_0(d) \in \down{(h_0(c) \cap h_0(a))}$, that is, $h_0(d) \subseteq h_0(c) \cap h_0(a)$, but if $(x, y) \in h_0(d)$, then $(x, y) \in h_0(c)$ since $h_0(d) \subseteq h_0(c)$. Therefore, every $(x, y) \in h_0(d)$ belongs to $h_0(c) \cap h_0(a)$, if $h_0(d)$ is a subset of $h_0(a)$.

    The second alternative with the assumption that $(x, y) \in h_0(a)$ is similar to the previous one.

    Suppose $h_0(d) \in (\down{h_0(c)} \cap \down{h_0(a)}) \vee (\down{h_0(c)} \cap \down{h_0(b)})$. So, we have either $h_0(d) \in \down{h_0(c)} \cap \down{h_0(a)}$ or $h_0(d) \in \down{h_0(c)} \cap \down{h_0(b)}$. So $h_0(d) \in \down{h_0(c)}$ anyway. If $h_0(d) \in \down{h_0(a)}$, then $d \leq a \leq a + b$, so $h_0(d) \in \down{h_0(a + b)}$.
    The second alternative is the same as the first one.

    So, $\operatorname{Id}(h_0(\mathcal{A}))$ is distributive, and, therefore, $\mathcal{A}_0$ is distibutive as a join-semilattice by \cite[Lemma 184]{gratzer2011lattice}.

    \item Let $g \in \mathcal{A}_0$ be a generator, suppose $g \leq a + b$, so there are $a' \leq a$ and $b' \leq b$ with $g = a' + b'$. $g$ is join-irreducible, thus, one has either $g = a'$ or $g = b'$. In fact, this statement generalises the fact that an element is join-prime iff it is join-irreducible in distributive lattices, see \cite[Lemma 5.11]{davey2002introduction}.
  \end{enumerate}
\end{proof}

The following lemma allows one to consider extensions finite elements of ${\bf R}(;,+)$ with bottoms as simple quantales. If $\mathcal{A}$ is finite and representable, then every $a \in \mathcal{A}_0$ has the form $\Sigma \{ g \in \mathcal{G}(\mathcal{A}) \: | \: g \leq a \}$, but every generator is join-prime by Proposition~\ref{distr}. So we have:

>>>>

\let\strokeL\L
\renewcommand\L{\mathbf{L}}

@@

\begin{lemma} \label{rep} Let $\mathcal{A} = \langle A, ;, + \rangle$ be a join semilattice-ordered semigroup,

  \begin{enumerate}
    \item If $\mathcal{A}$ is representable then $\exists$ has a winning strategy in $\mathcal{G}_{\omega}(\mathcal{A})$.
    \item If $|\mathcal{A}| = \omega$ and $\exists$ has a winning strategy in $\mathcal{G}_{\omega}(\mathcal{A})$ then $\mathcal{A}$ is representable.
  \end{enumerate}
\end{lemma}

@@

    \item
    Let $a, b \in \mathcal{A}$ such that $a \not\leq b$. We consider a play of the game where $\forall$ pick the pair $(a, b)$ in the initial round and plays $(\mathcal{N}, x, y, z, c, d)$ in the further rounds for each $x, y, z \in \operatorname{Nodes}(\mathcal{N})$ and $c,d \in \mathcal{A}$. Here, $c \in l_N(x, y)$ and $d \in l_N(y, z)$.

@@

\end{itemize}
  \end{enumerate}
\end{proof}

@@

\begin{proposition} \label{fin}
  Let $\mathcal{A}$ be a join semilattice-ordered semigroup, then there exists a game such that $\mathcal{A}$ is representable iff $\exists$ has a winning strategy in the all finite lenght games.
\end{proposition}

@@

We have already considered finite and countable algebras. Let us consider also arbitrary to complete the axiomatisation of ${\bf R}(;,+)$. We have the following statement.

\begin{lemma}
  Let $\mathcal{A}$ be a join-semilattice semigroup, then $\exists$ has a winning strategy has a winning strategy in $\mathcal{G}_m(\mathcal{A})$ iff $\mathcal{A}$ is elementary equivalent to some countable representable join-semilattice semigroups.
\end{lemma}

\begin{proof}
  The $\Leftarrow$ implication is obvious. If $\mathcal{A} \equiv \mathcal{A}'$, where $|\mathcal{A}'| = \omega$ and $\mathcal{A}' \in {\bf R}(;,+)$, then $\exists$ has a winning strategy in $\mathcal{G}_{\omega}(\mathcal{A}')$ (Lemma~\ref{rep}), so $\exists$ has a winning strategy in
  every $\mathcal{G}_{m}(\mathcal{A}')$ for $m < \omega$ (Proposition~\ref{fin}). Thus, so $\exists$ has a winning strategy in
  every $\mathcal{G}_{m}(\mathcal{A})$ since $\mathcal{A}$ and $\mathcal{A}'$ are elementary equivalent.

  Suppose $\exists$ has a winning strategy has a winning strategy in $\mathcal{G}_m(\mathcal{A})$ for $m < \omega$. The proof scheme is the same as for relation or cylindric algebras, so we provide the following sketch. In particular, we have the following proof that may be proved using \strokeL o\'s's Theorem similarly to.

  \begin{claim}
    If $\exists$ has a winning strategy has a winning strategy in $\mathcal{G}_m(\mathcal{A})$ for $m < \omega$, then $\exists$ has a winning strategy for $\mathcal{G}_{\omega}(\Pi_{U} \mathcal{A})$, where $\Pi_{U} \mathcal{A}$ is a non-principal ultrapower of $\mathcal{A}$ for $U$, an ultrafilter over $\omega$.
  \end{claim}

  Consider $\mathcal{B}$, a countable elementary equivalent subalgbera of $\Pi_{U} \mathcal{A}$ that exists by the downward L\"{o}wenheim-Skolem theorem.
  We construct the following elementary chain inxeded over $\omega$:
  \begin{center}
    $\mathcal{B}_0 \preceq \mathcal{B}_1 \preceq \dots \mathcal{B}_i \preceq\mathcal{B}_{i + 1} \preceq \dots (\preceq \Pi_{U} \mathcal{A})$.
  \end{center}
  Let us put $\mathcal{B}_0 := \mathcal{B}$.
  Suppose the initial segment $\mathcal{B}_0 \preceq \mathcal{B}_1 \preceq \dots \mathcal{B}_i$ has been already constructed. Let us constuct $\mathcal{B}_{i + 1}$.

  Consider a function $\sigma$ that takes an initial game of the play $\mathcal{G}_{\omega}(\Pi_{U} \mathcal{A})$, say, $\mathcal{N}_0 \subseteq \mathcal{N}_1 \subseteq \dots \mathcal{N}_k$ (all $k$) and Abelard's move and returns $\exists$'s repsonse, $\mathcal{N}_{k + 1}$, an extension of $\mathcal{N}_k$.

  Consider $\sigma \upharpoonright_{\mathcal{B}_i}$, a restriction of $\sigma$, where $\forall$ takes elements only from $\mathcal{B}_i$ for his moves. The range of this restriction consists of $\Pi_{U} \mathcal{A}$-networks, and $E$, the set of elements that occur in those networks is also countable. We define $\mathcal{B}_{i + 1}$ as a countable elementary subalgebra of $\Pi_{U} \mathcal{A}$ that contains $\mathcal{B}_{i}$ and $E$. The existence follows from the downward L\"{o}wenheim-Skolem theorem.

  Let us put $\mathfrak{B} = \bigcup \limits_{i < \omega} \mathcal{B}_i$. Note that $\mathfrak{B}$ is countable. By Tarksi-Vaught elementary chain union theorem, for each $i < \omega$ $\mathcal{B}_i \preceq \mathfrak{B}$. Therefore, $\mathfrak{B} \equiv \Pi_{U} \mathcal{A}$. On the other hand, $\mathcal{A} \equiv \Pi_{U} \mathcal{A}$.
  Moreover, $\mathfrak{B}$ is representable and $\mathfrak{B} \equiv \mathcal{A}$.
\end{proof}

@@

TODO: complete the argument
